Supplementary material: When does vapor pressure deficit drive or reduce evapotranspiration?

### 2019-03-01 doc/paper/body.tex
done through methods, onto results

\section{Methods}
\label{methods}
The Penman-Monteith equation \citep [hereafter PM,][]{Penman_1948,
  Monteith_1965} estimates ET as a function of observable atmospheric
variables and surface conductances:
% \begin{linenomath*}
  \begin{equation}
    \label{orig_pen}
    ET = \frac{\Delta R_{net} + g_a \rho_a c_p VPD}{\Delta + \gamma(1 + \frac{g_a}{g_s})},
  \end{equation}
% \end{linenomath*}
where $\Delta$ is the change in saturation vapor pressure with
temperature, given by Clausius-Clapeyron ($\frac{d \; e_s}{d \; T}$),
$R_{net}$ is the net radiation minus ground heat flux, $g_a$ is
aerodynamic conductance, $\rho_a$ is air density, $c_p$ is specific
heat of air at constant pressure, $\gamma$ is the psychometric
constant, and $g_s$ is the stomatal conductance (Table
\ref{definitions}). The issue with PM is that $g_s$ is a function of
carbon uptake, which is has a strong functionally relation to ET
through stomatal function. Therefore, when PM is formulated in terms
of $g_s$, it is an implicit funciton of ET as a function of itself
rather than an explicit function of ET. Here we will derive a new form
of PM in which ET is an explicit function of plant parameters and
environmental conditions, and use it to assess the ecosystem scale
response to VPD (by taking a partial derivative).

\begin{table}
  \caption{Definition of symbols and variables, with citation for how
    values are calculated, if applicable.}
  \label{definitions}
  \centering \footnotesize
  \begin{tabular}{l c c c}
    \hline
    Variable & Description & Units & Citation \\
    \hline
    $e_s$  & saturation vapor pressure & Pa  & - \\
    $T$  & temperature  & K & - \\
    $P$  & pressure & Pa  & - \\
    $\Delta$  & $\frac{\partial e_s}{\partial T}$ & Pa K$^{-1}$ & - \\
    $R_{net}$  & net radiation at land surface minus ground heat flux & W m$^{-2}$   & - \\
    $g_a$  & aerodynamic conductance & m s$^{-1}$  & \citet{Shuttleworth_2012} \\
    $\rho_a$  & air density & kg m$^{-3}$  & - \\
    $c_p$  & specific heat capacity of air at constant pressure & J K$^{-1}$ kg$^{-1}$ & - \\
    $VPD$  & vapor pressure deficit & Pa  & - \\
    $\gamma$  & psychometric constant & Pa K$^{-1}$   & - \\
    $g_{0}$  & leaf-scale mimimum stomatal conductance & m s$^{-1}$  & \citet{MEDLYN_2011} \\

    $g_{s-leaf}$  & leaf-scale stomatal conductance & m s$^{-1}$  & \citet{MEDLYN_2011} \\
    $g_{s}$  &  stomatal conductance & m s$^{-1}$
                                   & \citet{Medlyn_2017} \\
    $g_{1-leaf}$  & leaf-scale slope parameter & Pa$^{0.5}$
                                   & \citet{MEDLYN_2011} \\
    $g_{1}$  & ecosystem-scale slope parameter & Pa$^{0.5}$ & \citet{Medlyn_2017} \\
    $R$ & universal gas constant & J mol$^{-1}$ K$^{-1}$ & - \\
    $R_{air}$ & gas constant of air & J  K$^{-1}$ kg$^{-1}$ & - \\
    $\sigma$ & uncertainty parameter & -& - \\
    $c_a$ & CO$_2$ concentration & $\mu$ mol CO$_2$ mol$^{-1}$ air& - \\
    $\lambda = \frac{\partial \; transpiration}{\partial\; CO_2\; assimilation}$ & marginal water cost of leaf carbon & mol H$_2$O mol$^{-1}$ CO$_2$ & - \\
    $\Gamma$ & CO$_2$ compensation point & - & - \\
    $\Gamma^*$ & CO$_2$ compensation point without dark respiration & - & - \\
    \hline
  \end{tabular}
\end{table}

\citet{MEDLYN_2011} developed a model for stomatal conductance ($g_s$)
by combining an optimal photosynthesis theory \citep{Cowan_1977} with an empirical approach, which describes the
dependence of $g_s$ to VPD. This resulted in the following model for
leaf-scale stomatal conductance:

% \begin{linenomath*}
  \begin{equation}
    g_{s-leaf} = g_0 + 1.6 \left(1 +
      \frac{g_{1-leaf}}{\sqrt{VPD}}\right) \frac{A}{c_a},
    \label{leaf_medlyn}
  \end{equation}
% \end{linenomath*}
where $g_{1-leaf}$ is a leaf-scale ``slope'' parameter, A is the net
CO$_2$ assimilation rate, and $c_a$ is the atmospheric CO$_2$
concentration at the leaf surface. \cite{MEDLYN_2011} relate the slope
parameter ($g_{1-leaf}$) to physical parameters as:
% \begin{linenomath*}
  \begin{equation}
    g_{1-leaf} = \sqrt{\frac{3 \, \Gamma^* \,
        \lambda}{1.6}},\footnote{Note this expression has units of of
      (mmol mol$^{-1}$)$^{1/2}$, but this can be converted to
      Pa$^{1/2}$ using the ideal gas law.}
    \label{slope}
  \end{equation}
% \end{linenomath*}

where $\Gamma^*$ is the CO$_2$ compensation point for photosynthesis
(without dark respiration), and $\lambda$ is the marginal water cost
of leaf carbon
($\frac{\partial \; \text{transpiration}}{\partial \; A}$)
\citep{Farquhar_1980, Katul_2009}. So,
$g_{1-leaf}$ is a leaf-scale term reflecting the trade-off of water for
carbon uptake. The higher $g_{1-leaf}$, the more open the stomata and
the more they release water in exchange for carbon.

The Medlyn model adopted to ecosystem scale analysis in
\citet{Medlyn_2017}. In units of m s$^{-1}$, the ecosystem scale
stomatal conductance is:

@@

  where GPP is the ecosystem scale gross primary production, and $g_1$
  is an ecosystem scale analogue to $g_{1-leaf}$.

  While Eq. (\ref{medlyn}) can be used in PM (Eq. (1)), it will make
  analytical work with the function intractable because $GPP$ is
  functionally related to ET itself. Additionally, a perturbation to
  VPD should induce a physiological plant response that will alter GPP
  and cause an indirect change in stomatal conductance, in addition to
  the direct effect of VPD in Eq. (\ref{medlyn}). Therefore, in order
  to derive the response of ET to VPD, we must account for the
  functional relationship between GPP, ET, and VPD, and its effect on
  stomatal conductance. We can use aforementioned semi-empirical
  results of \citet{Zhou_2015}, which were inspired optimal
  photosynthesis theory, as a tool to approach this
  problem. \citet{Zhou_2015}, showed that underlying Water Use
  Efficiency (uWUE):

@@

  is relatively constant across time within a plant functional type,
  and correctly captures a constant relationship between GPP, ET and
  VPD over a diurnal cycle during the growing season
  \citep{Zhou_2014}. The theoretical derivation of the square root VPD
  dependence in $uWUE$ leverages the same assumptions used in
  \cite{MEDLYN_2011} to derive the square-root VPD dependence of the
  stomatal conductance model (Eq.  (\ref{medlyn})).  We can use uWUE
  to remove the $GPP$ dependence from $g_s$ in a way that makes PM
  analytically tractable:

@@

where $\Gamma$ is the CO$_2$ compensation point \citep[Eq. (5)
in][]{Zhou_2014}. So uWUE is related to atmospheric CO$_2$
concentration and compensation point, and is inversely proportional to
the marginal water cost of leaf carbon.

@@

  deriving an analytical framework for ecosystem response to
  atmospheric demand with environmental conditions held
  fixed. There are a few subtleties to taking the derivative in
  Eq. (\ref{d_et}): $\Delta$ ($\frac{d e_{s}}{d T}$) and $VPD$ are
  functionally related, so while taking the derivative we evaluate
  $\frac{\partial \; ET}{\partial \; VPD} = \frac{\partial \; ET}
  {\partial \; e_s} \frac{\partial \; e_s}{\partial \; VPD}
  \Big|_{\text{RH fixed}} + \frac{\partial \; ET}{\partial \; RH}
  \frac{\partial \; RH}{\partial \; VPD} \Big|_{\text{$e_s$
      fixed}}$. $RH$ and $e_s$ are assumed to be approximately
  independent, which is supported by data (not shown).

  This derivation relied either implicitly or explicitly on several
  assumptions. First, we assume that VPD at the leaf surface is the
  same as VPD at measurement height; physically this implies that
  leaves are perfectly coupled to the atmosphere. In reality, for some
  conditions and plant types the leaves can become decoupled from the
  boundary layer \citep{De_2017, Medlyn_2017}. Therefore, our
  derivation will be most applicable in times like the growing season
  (when we also expect uWUE to be most valid), when relatively high
  insolation induces instability and convective boundary layers, and
  we would expect the surface to be generally well coupled. An
  additional assumption in the formulation of $uWUE$ \citep{Zhou_2014,
    Zhou_2015} and \citet{Medlyn_2017}'s stomatal conductance model is
  that direct soil evaporation contributions to ET remain small
  relative to transpiration. Again, this should be more true during
  the growing season. The ratio of evaporation to transpiration may
  increase immediately after rainfall events due to high soil moisture
  and ponding, but VPD is generally low anyways during these
  times. However, some plant types allow for systematically larger
  contributions of evaporation in ET, particularly those with sparse
  canopies and smaller relative amounts of transpiration. We therefore
  might expect that the theory will be most applicable to forest PFTs,
  which will be most strongly coupled to the boundary layer due to
  larger surface roughness, and will also generally have the highest
  ratios of transpiration to evaporation. Finally, we assume that
  $g_1$ and $uWUE$ are constant with respect to the conceptual VPD
  perturbation. Both quantities have been shown to be relatively
  constant with respect to changes in VPD \citep{Franks_2017,
    Zhou_2014}. These parameters will however vary with plant species
  and characteristics \citep[e.g. wood density, ][]{Lin_2015}, as well
  as environmental conditions including soil water content
  \citep{Manzoni2013} and temperature.

  As Eq. (\ref{d_et}) is a partial derivative, it's utility is as a
  conceptual model for the change in ET in response to VPD with all
  other variables held fixed. This concept is a useful tool for
  identifying the effect of VPD on ET through atmospheric demand and
  plant response, and allows a practitioner to disentangle complicated
  feedbacks when many quantities co-vary. However users of
  Eq. (\ref{d_et}) should take care that their interpretation matches
  the assumptions inherent in a partial derivative. For example,
  results will only be as valid as the assumption that g$_1$ and uWUE
  (and by extension $\lambda$) are fixed with respect to the user's
  conceptual change in VPD. Fortunately, these quantities are
  relatively constant for many under many scenarios
  \citep{Franks_2017}. Care must also be taken with possible indirect
  effects associated with a change in VPD: for example, a change in ET
  induced by a change in VPD can cause a change in surface
  temperature, which would drive a change in net radiation. These
  types of indirect effects and feedbacks are not considered in
  Eq. (\ref{d_et}): temperature (a variable) is mathematically fixed.

@@

  ** comment on temperature and acclimation??? **
  ** impact of T is much smaller than VPD? **
  ** paragraph on soil moisture **

@@

So far we have discussed the role of different parameters in
modulating the ET response. However, the form of the ET response is
imposed by our choice of stomatal conductance model. We now examine
how the stomatal model choice, and specifically the exponent of the
VPD dependence in the stomatal conductance model, can impact the
general form of the ET-VPD relationship. There is a theoretical basis
for the square root VPD dependence in both the stomatal conductance
model and $uWUE$ based on the assumption that stomata behave to
maximize carbon gain while minimizing water loss, which observations
also generally support \citep{Lloyd_1991, MEDLYN_2011, Lin_2015,
  Zhou_2014, Zhou_2015, Medlyn_2017}. However, some purely empirical
results that fit the exponent of the VPD dependence to data have shown
that it may vary slightly from 1/2, suggesting that stomata, as well
as ecosystem-scale quantities based on stomata theory, may not always
function optimally \citep{Zhou_2015, Lin_2018}. Specifically with
regards to $uWUE$, one would not expect that this ecosystem scale WUE
quantity will respond to VPD exactly analogously to stomata. Direct
soil evaporation's contributions to ET should shift the exponent of
the VPD dependence, especially at conditions of low GPP when we would
expect a systematically larger portion of direct soil evaporation
contributions to ET, because we would also expect lower amounts of
transpiration at low GPP. \citet{Zhou_2015}'s results corroborate
this: they found a mean empirically fit exponential VPD dependence of
0.55, varying slightly from the theoretically optimal value of of 0.5
for AmeriFlux sites. \citet{Lin_2018}'s results also show variance in
the empirical exponent of the VPD dependence of the stomatal
conductance model, but interpretation of this variance is more
difficult as \citet{Lin_2018} do not handle GPP/A dependence of
stomatal conductance in a directly analogous manner to the optimal
theory in \citet{MEDLYN_2011} and \citet{Medlyn_2017}. Regardless,
given that these recent results on the relationship between VPD, GPP,
and ET \citep{MEDLYN_2011, Zhou_2014, Zhou_2015, Medlyn_2017} form the
backbone of our analysis and are what allowed us to derive an explicit
ET expression for the first time (Eq. (\ref{et})), we will analyze if
and how assumptions about the exponent of the VPD dependence impacts
the shape of the ET-VPD dependence. This analysis is also important to
understand whether the choice of stomatal conductance model alters the
fundamental behavior of the ET-VPD relationships, as many commonly
used models utilize a VPD exponent other than the 1/2 suggested by
optimal theory \citep[e.g. ][ which uses an exponent of
1]{Leuning_1990}. It is useful to understand how the choice of model
imposes the functional form of the ET-VPD relationship, given its
fundamental importance for land-atmosphere interactions.

@@

The results presented in Sect. \ref{results} indicates that for a
given $uWUE$ and $g_1$, the ET dependence on VPD should be concave
upward. In other words, there should be some local minimum in ET at a
critical VPD$_{crit}$, assuming the scaling and plant terms
(e.g. aerodynamic conductance, $\Delta$, $g_1$ and $uWUE$) are held
fixed. This result warrants further investigation, because to our
knowledge no earlier work has derived the theoretical ecosystem-scale
relationship between ET and VPD while controlling for other
environmental conditions. In particular there is an apparent lack of
consensus over whether the shape of the ET-VPD curve should be concave
upward (our result) or concave downward in the absence of dramatic
water stress. Given that understanding the ET-VPD relationship of the
one-way plant response is fundamental to hypothesizing about any
feedbacks between the land surface and the atmosphere, we analyze why
our derived ET-VPD relationship is concave upward, particularly with
respect to the exponent of VPD dependence in $uWUE$ and the Medlyn
unified stomatal conductance model, as other models and empiracle
results have suggested exponents varying from 1/2 \citep{Leuning_1990,
  Zhou_2015, Lin_2018}.

@@

\begin{itemize}
  \item For optimal behavior (n, m = 1/2) the ET-VPD curve will be
    concave up regardless of the magnitude of the plant constants
    $g_1$ and $uWUE$. Therefore, the general concave up nature of our
    results, given an assumption of optimal behavior, is insensitive
    to plant type.
  \item For all physically possible exponents of VPD dependence ($n,
    m$), whether the solution is concave up or concave down does not
    depend on $uWUE$.
  \item In general, increasing the exponent of VPD dependence for
    either $uWUE*$ or $g_*$ increases the likelihood of a concave down
    result. Additionally, as the exponent of VPD dependence increases
    from the optimum value of 1/2, whether the curve is concave upward
    or concave downward becomes a function of the plant specific slope
    parameter $g_*$, through non-dimensional VPD
    ($VPD^m/g_*$). Because the exponent of the VPD dependencies is
    capable of altering the fundamental shape of ET-VPD dependence,
    future research investment in understanding the exact VPD
    dependence of stomatal conductance, and further reconciliation of
    empirical and theoretical stomatal and ecosystem behavior should
    be prioritized.
\end{itemize}

>>>>

\section{Methods}
\label{methods}
The Penman-Monteith equation \citep [hereafter PM,][]{Penman_1948,
  Monteith_1965} estimates ET as a function of observable atmospheric
variables and surface conductances:
% \begin{linenomath*}
  \begin{equation}
    \label{orig_pen}
    ET = \frac{\Delta R_{net} + g_a \rho_a c_p VPD}{\Delta + \gamma(1 + \frac{g_a}{g_s})},
  \end{equation}
% \end{linenomath*}
  where $\Delta$ is the change in saturation vapor pressure with
  temperature, given by Clausius-Clapeyron
  ($\frac{d \; e_s}{d \; T}$), $R_{net}$ is the net radiation minus
  ground heat flux, $g_a$ is aerodynamic conductance, $\rho_a$ is air
  density, $c_p$ is specific heat of air at constant pressure,
  $\gamma$ is the psychometric constant, and $g_s$ is the stomatal
  conductance (Table \ref{definitions}). The issue with PM is that
  $g_s$ is a function of carbon uptake, which is has a strong
  functionally relation to ET through stomatal function. Therefore,
  when PM is formulated in terms of $g_s$, it is an implicit function
  of ET itself rather than an explicit function of ET. Here we will
  derive a new form of PM in which ET is an explicit function of plant
  parameters and environmental conditions, and use it to assess the
  ecosystem scale response to VPD (by taking a partial derivative).

\begin{table}
  \caption{Definition of symbols and variables, with citation for how
    values are calculated, if applicable.}
  \label{definitions}
  \centering \footnotesize
  \begin{tabular}{l c c c}
    \hline
    Variable & Description & Units & Citation \\
    \hline
    $e_s$  & saturation vapor pressure & Pa  & - \\
    $T$  & temperature  & K & - \\
    $P$  & pressure & Pa  & - \\
    $\Delta$  & $\frac{\partial e_s}{\partial T}$ & Pa K$^{-1}$ & - \\
    $R_{net}$  & net radiation at land surface minus ground heat flux & W m$^{-2}$   & - \\
    $g_a$  & aerodynamic conductance & m s$^{-1}$  & \citet{Shuttleworth_2012} \\
    $\rho_a$  & air density & kg m$^{-3}$  & - \\
    $c_p$  & specific heat capacity of air at constant pressure & J K$^{-1}$ kg$^{-1}$ & - \\
    $VPD$  & vapor pressure deficit & Pa  & - \\
    $\gamma$  & psychometric constant & Pa K$^{-1}$   & - \\
    $g_{s}$  &  stomatal conductance & m s$^{-1}$
                                   & \citet{Medlyn_2017} \\
    $g_{1}$  & ecosystem-scale slope parameter & Pa$^{0.5}$ & \citet{Medlyn_2017} \\
    $R$ & universal gas constant & J mol$^{-1}$ K$^{-1}$ & - \\
    $R_{air}$ & gas constant of air & J  K$^{-1}$ kg$^{-1}$ & - \\
    $\sigma$ & uncertainty parameter & -& - \\
    $c_a$ & CO$_2$ concentration & $\mu$ mol CO$_2$ mol$^{-1}$ air& - \\
    $\lambda = \frac{\partial \; transpiration}{\partial\; CO_2\; assimilation}$ & marginal water cost of leaf carbon & mol H$_2$O mol$^{-1}$ CO$_2$ & - \\
    $\Gamma$ & CO$_2$ compensation point & - & - \\
    $\Gamma^*$ & CO$_2$ compensation point without dark respiration &
                                                                      - & - \\
    GPP & gross primary production & $\mu$-mol C s$^{-1}$ m$^{-2}$ & -
    \\
    ET & evapotranspiration & W m$^{-2}$ & - \\
   uWUE & underlying water use efficiency & $\mu$-mol C Pa$^{0.5}$
                                            J$^{-1}$ ET  &
                                                           \citet{Zhou_2015} \\
    \hline
  \end{tabular}
\end{table}

\citet{MEDLYN_2011} developed a model for stomatal conductance ($g_s$)
by combining an optimal photosynthesis theory \citep{Cowan_1977} with
an empirical approach, which describes the dependence of $g_s$ to
VPD. They also extended this model to the ecosystem scale
\citep{Medlyn_2017}:

@@

  where $g_{1}$ is a VPD ``slope'' parameter, GPP is the ecosystem
  scale gross primary production, and $c_a$ is the atmospheric CO$_2$
  concentration at the canopy. \cite{MEDLYN_2011} relate the slope
  parameter ($g_{1}$) to physical parameters as:
% \begin{linenomath*}

  \begin{equation}
    g_{1}  \propto \sqrt{\frac{3 \, \Gamma^* \,
        \lambda}{1.6}},\footnote{Note this expression has units of of
      (mmol mol$^{-1}$)$^{1/2}$, but this can be converted to
      Pa$^{1/2}$ using the ideal gas law.}
    \label{slope}
  \end{equation}
% \end{linenomath*}

where $\Gamma^*$ is the CO$_2$ compensation point for photosynthesis
(without dark respiration), and $\lambda$ is the marginal water cost
of leaf carbon
($\frac{\partial \; \text{transpiration}}{\partial \; CO_2 assimilation}$)
\citep{Farquhar_1980, Katul_2009}. So,
$g_{1}$ is a leaf-scale term reflecting the trade-off of water for
carbon uptake. The higher $g_{1}$, the more open the stomata and
the more they release water in exchange for carbon.

While Eq. (\ref{medlyn}) can be used in PM (Eq. (1)), it will make
analytical work with the function intractable because GPP is
functionally related to ET itself. Additionally, a perturbation to VPD
should induce a physiological plant response that will alter GPP and
cause an indirect change in stomatal conductance, in addition to the
direct effect of VPD in Eq. (\ref{medlyn}). Therefore, in order to
derive the full plant response of ET to VPD, we must account for the
functional relationship between GPP, ET, and VPD, and its effect on
stomatal conductance. We can use aforementioned semi-empirical results
of \citet{Zhou_2015}, which were inspired optimal photosynthesis
theory, as a tool to approach this problem. \citet{Zhou_2015}, showed
that underlying Water Use Efficiency (uWUE):

@@

  is relatively constant across time within a plant functional type,
  and correctly captures a constant relationship between GPP, ET and
  VPD over a diurnal cycle during the growing season
  \citep{Zhou_2014}. The theoretical derivation of the square root VPD
  dependence in uWUE leverages the same assumptions used in
  \cite{MEDLYN_2011} to derive the square-root VPD dependence of the
  stomatal conductance model (Eq.  (\ref{medlyn})).  We can use uWUE
  to remove the $GPP$ dependence from $g_s$ in a way that makes PM
  analytically tractable:

@@

where $\Gamma$ is the CO$_2$ compensation point \citep[Eq. (5)
in][]{Zhou_2014}. So uWUE is related to atmospheric CO$_2$
concentration and compensation point, and is inversely proportional to
the marginal water cost of leaf carbon ($\lambda$). This relationship
with $\lambda$ ($\propto \lambda^{-1/2}$) is important as it is the inverse of g$_1$'s
relationship with $\lambda$ ($\propto \sqrt{\lambda}$). So, uWUE
should increase as $\lambda$ decreases, and g$_1$ should decrease as
$\lambda$ decreases.

@@

  providing analytical framework for ecosystem response to
  atmospheric demand with environmental conditions held
  fixed. There are a few subtleties to taking the derivative in
  Eq. (\ref{d_et}): $\Delta$ ($\frac{d e_{s}}{d T}$) and $VPD$ are
  functionally related, so while taking the derivative we evaluate
  $\frac{\partial \; ET}{\partial \; VPD} = \frac{\partial \; ET}
  {\partial \; e_s} \frac{\partial \; e_s}{\partial \; VPD}
  \Big|_{\text{RH fixed}} + \frac{\partial \; ET}{\partial \; RH}
  \frac{\partial \; RH}{\partial \; VPD} \Big|_{\text{$e_s$
      fixed}}$. $RH$ and $e_s$ are assumed to be approximately
  independent, which is supported by data (not shown).

  This derivation relied either implicitly or explicitly on several
  assumptions. First, we assume that VPD at the leaf surface is the
  same as VPD at measurement height; physically this implies that
  leaves are perfectly coupled to the atmosphere. In reality, for some
  conditions and plant types the leaves can become decoupled from the
  boundary layer \citep{De_2017, Medlyn_2017}. Therefore, our
  derivation will be most applicable in times like the growing season
  (when we also expect uWUE to be most valid), when relatively high
  insolation induces instability and convective boundary layers, and
  we would expect the surface to be generally well coupled. An
  additional assumption in the formulation of uWUE \citep{Zhou_2014,
    Zhou_2015} and \citet{Medlyn_2017}'s stomatal conductance model is
  that direct soil evaporation contributions to ET remain small
  relative to transpiration. Again, this should be more true during
  the growing season. The ratio of evaporation to transpiration may
  increase immediately after rainfall events due to high soil moisture
  and ponding, but VPD is generally low anyways during these
  times. However, some plant types allow for systematically larger
  contributions of evaporation in ET, particularly those with sparse
  canopies and smaller relative amounts of transpiration. We therefore
  might expect that the theory will be most applicable to forest PFTs,
  which will be most strongly coupled to the boundary layer due to
  larger surface roughness, and will also generally have the highest
  ratios of transpiration to evaporation. Finally, we assume that
  $g_1$ and uWUE are constant with respect to the conceptual VPD
  perturbation. Both quantities have been shown to be relatively
  constant with respect to changes in VPD \citep{Franks_2017,
    Zhou_2014}. These parameters will however vary with plant species
  and characteristics \citep[e.g. wood density, ][]{Lin_2015}, as well
  as environmental conditions including soil water content and
  temperature \citep{Lin_2015, Manzoni2013}. Exploring possible soil
  moisture (in)dependnce of the plant parameters (g$_1$ and uWUE) is
  particularly interesting because soil moisture only enters the
  partial derivative directly through these plant prameters. If the
  plant parameters are weak functions of soil moisture then the theory
  can be directly applied to a broader range of conceptual VPD
  scenarios, including observed compound events between high VPD and
  low soil moisture \citep{Zhou_2019}. Supplementary material for this
  manuscript explores the soil moisture dependence of the plant
  parameters, but excessive noise and inconsistencies preclude
  conclusions about the nature of any dependence. We provide it in
  case it is of interest to the reader, and to motivate future
  research.

   Because Eq. (\ref{d_et}) is a partial derivative, it's utility is as a
   conceptual model for the change in ET in response to VPD with all
   other variables held fixed. This provides is a useful tool for
   identifying the effect of VPD on ET through atmospheric demand and
   plant response, and allows a practitioner to disentangle
   complicated feedbacks when many quantities co-vary. However users
   of Eq. (\ref{d_et}) should take care that their interpretation
   matches the assumptions inherent in a partial derivative. For
   example, results will only be as valid as the assumption that g$_1$
   and uWUE (and by extension $\lambda$) are fixed with respect to the
   user's conceptual change in VPD. Care must also be taken with
   possible indirect effects associated with a change in VPD: for
   example, a change in ET induced by a change in VPD can cause a
   change in surface temperature, which would drive a change in net
   radiation. These types of indirect effects and feedbacks are not
   considered in Eq. (\ref{d_et}): temperature (a variable) is
   mathematically fixed.

\subsection{Framing Eq. \ref{d_et} with previous research }

@@

@@

So far we have discussed the role of different parameters in
modulating the ET response. However, the form of the ET response is
imposed by our choice of stomatal conductance model. We now examine
how the stomatal model choice, and specifically the exponent of the
VPD dependence in the stomatal conductance model, can impact the
general form of the ET-VPD relationship. There is a theoretical basis
for the square root VPD dependence in both the stomatal conductance
model and uWUE based on the assumption that stomata behave to
maximize carbon gain while minimizing water loss, which observations
also generally support \citep{Lloyd_1991, MEDLYN_2011, Lin_2015,
  Zhou_2014, Zhou_2015, Medlyn_2017}. However, some purely empirical
results that fit the exponent of the VPD dependence to data have shown
that it may vary slightly from 1/2, suggesting that stomata, as well
as ecosystem-scale quantities based on stomata theory, may not always
function optimally \citep{Zhou_2015, Lin_2018}. Specifically with
regards to uWUE, one would not expect that this ecosystem scale WUE
quantity will respond to VPD exactly analogously to stomata. Direct
soil evaporation's contributions to ET should shift the exponent of
the VPD dependence, especially at conditions of low GPP when we would
expect a systematically larger portion of direct soil evaporation
contributions to ET, because we would also expect lower amounts of
transpiration at low GPP. \citet{Zhou_2015}'s results corroborate
this: they found a mean empirically fit exponential VPD dependence of
0.55, varying slightly from the theoretically optimal value of of 0.5
for AmeriFlux sites. \citet{Lin_2018}'s results also show variance in
the empirical exponent of the VPD dependence of the stomatal
conductance model, but interpretation of this variance is more
difficult as \citet{Lin_2018} do not handle GPP/A dependence of
stomatal conductance in a directly analogous manner to the optimal
theory in \citet{MEDLYN_2011} and \citet{Medlyn_2017}. Regardless,
given that these recent results on the relationship between VPD, GPP,
and ET \citep{MEDLYN_2011, Zhou_2014, Zhou_2015, Medlyn_2017} form the
backbone of our analysis and are what allowed us to derive an explicit
ET expression for the first time (Eq. (\ref{et})), we will analyze if
and how assumptions about the exponent of the VPD dependence impacts
the shape of the ET-VPD dependence. This analysis is also important to
understand whether the choice of stomatal conductance model alters the
fundamental behavior of the ET-VPD relationships, as many commonly
used models utilize a VPD exponent other than the 1/2 suggested by
optimal theory \citep[e.g. ][ which uses an exponent of
1]{Leuning_1990}. It is useful to understand how the choice of model
imposes the functional form of the ET-VPD relationship, given its
fundamental importance for land-atmosphere interactions.

@@

The results presented in Sect. \ref{results} indicates that for a
given uWUE and $g_1$, the ET dependence on VPD should be concave
upward. In other words, there should be some local minimum in ET at a
critical VPD$_{crit}$, assuming the scaling and plant terms
(e.g. aerodynamic conductance, $\Delta$, $g_1$ and uWUE) are held
fixed. This result warrants further investigation, because to our
knowledge no earlier work has derived the theoretical ecosystem-scale
relationship between ET and VPD while controlling for other
environmental conditions. In particular there is an apparent lack of
consensus over whether the shape of the ET-VPD curve should be concave
upward (our result) or concave downward in the absence of dramatic
water stress. Given that understanding the ET-VPD relationship of the
one-way plant response is fundamental to hypothesizing about any
feedbacks between the land surface and the atmosphere, we analyze why
our derived ET-VPD relationship is concave upward, particularly with
respect to the exponent of VPD dependence in uWUE and the Medlyn
unified stomatal conductance model, as other models and empiracle
results have suggested exponents varying from 1/2 \citep{Leuning_1990,
  Zhou_2015, Lin_2018}.

@@

\begin{itemize}
  \item For optimal behavior (n, m = 1/2) the ET-VPD curve will be
    concave up regardless of the magnitude of the plant constants
    $g_1$ and uWUE. Therefore, the general concave up nature of our
    results, given an assumption of optimal behavior, is insensitive
    to plant type.
  \item For all physically possible exponents of VPD dependence ($n,
    m$), whether the solution is concave up or concave down does not
    depend on uWUE.
  \item In general, increasing the exponent of VPD dependence for
    either $uWUE*$ or $g_*$ increases the likelihood of a concave down
    result. Additionally, as the exponent of VPD dependence increases
    from the optimum value of 1/2, whether the curve is concave upward
    or concave downward becomes a function of the plant specific slope
    parameter $g_*$, through non-dimensional VPD
    ($VPD^m/g_*$). Because the exponent of the VPD dependencies is
    capable of altering the fundamental shape of ET-VPD dependence,
    future research investment in understanding the exact VPD
    dependence of stomatal conductance, and further reconciliation of
    empirical and theoretical stomatal and ecosystem behavior should
    be prioritized.
\end{itemize}

### 2019-03-01 doc/paper/hess-supplement.tex
done through methods, onto results

>>>>

\section{Description of data}
\subsection{Data}
\label{data}
We use both meteorological and eddy-covariance data from the FLUXNET2015
database (data available at \sloppy
\url{https://fluxnet.fluxdata.org/data/fluxnet2015-dataset/} \sloppy),
including all Tier 1 sites with at least four years of data and
observations of the variables described in the methods section (Sect.
\ref{methods}). Sixty-six sites met these requirements, and were grouped
into nine plant functional types (PFT) according to the International
Geosphere-Biosphere Programme vegetation classification scheme
\citep{Loveland_1999}: cropland (CRO), grass (GRA), deciduous broadleaf
forest (DBF), evergreen broadleaf forest (EBF), evergreen needleleaf
forest (ENF), mixed forest (MF), closed shrub (CSH), savannah (SAV), and
woody savannah (WSA) (site locations and citations are in Fig.
\ref{map_fig} and Appendix \ref{flux_sites}, respectively).

\begin{figure}[h]
  \centering \includegraphics[trim={0 3cm 0 3cm}, clip]{./map.pdf}
  \caption{Plant functional type and location of FLUXNET2015 sites
    used in this analysis.}
  \label{map_fig}
\end{figure}
We filter and quality control the FLUXNET-2015 data using a similar
procedure as \cite{Zhou_2015}:

\begin{itemize}
\item Only measured or highest (``good'') quality gapfilled data,
  according to quality control flags, are used.
\item To isolate the growing season, we only use days in which the
  average Gross Primary Productivity (GPP) exceeds 10\% of the
  observed 95th percentile of GPP for a given site. GPP is calculated
  using the nighttime respiration partitioning method.
\item We remove days with rain and the day following to avoid issues
  with rain interception and sensor saturation at high relative
  humidity (\cite{Medlyn_2017}).
\end{itemize}
Additionally, as in \citet{Lin_2018}, we restrict data to the daytime,
which is identified when downwelling shortwave radiation is greater
than 50 W m$^{-2}$ and sensible heat flux is greater than 5 W
m$^{-2}$. To reduce the chance of sensor saturation at high relative
humidity, we remove all time steps for which VPD is less than .01 kPa,
and to reduce errors at low windspeeds we remove all periods with wind
magnitudes less than 0.5 m s$^{-1}$. Timesteps with negative observed
GPP or ET are also removed, and we aggregate half hourly data to
hourly averages to reduce noise \citep{Lin_2018}. After these quality
control procedures, 400,983 upscaled hourly observations remain.
